%%% TokenSmith 3-minute class presentation
%%% Compile with pdflatex (Beamer). Aspect ratio 16:9.

\documentclass[aspectratio=169,11pt]{beamer}

\usetheme{Madrid}
\usecolortheme{seahorse}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}[frame number]

\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta}
\usepackage{booktabs}

\title[TokenSmith on Consumer GPUs]{Optimizing TokenSmith for Consumer-GPU Hardware}
\subtitle{A Local-First RAG Pipeline}
\author{Shivam Singhal}
\institute{Georgia Tech}
\date{CS 6423, Spring 2026}

\begin{document}

%% ----------------------------------------------------------
\begin{frame}
  \titlepage
\end{frame}

%% ----------------------------------------------------------
\begin{frame}{The problem}
TokenSmith is a local-first RAG system for textbook question answering.
No external API calls, everything runs on the student's machine.

\bigskip

The original system works, but the cold path is brutal:
\begin{itemize}
  \item Sequential PDF extraction: \textbf{9{,}650\,s} (about 2h 40m)
  \item 4\,B embedder build: another 5 to 7 minutes on a server GPU
  \item First answer: a few seconds, after all of that
\end{itemize}

\bigskip

\begin{block}{}
\centering
\textbf{Cold time-to-first-token on a laptop: $\sim$\,2h 45m.}\\
Nobody waits that long for one textbook.
\end{block}
\end{frame}

%% ----------------------------------------------------------
\begin{frame}{The pipeline and where we attacked it}
\centering
\begin{tikzpicture}[
  > = {Stealth[length=2mm]},
  thick,
  font=\scriptsize,
  stage/.style={draw, rounded corners=2pt, fill=blue!10,
                minimum width=14mm, minimum height=7mm, align=center},
  io/.style={font=\footnotesize\itshape},
  arr/.style={->, thick}
]
  % Cold ingestion row
  \node[io]    (pdf) at (0, 0)   {PDF};
  \node[stage] (ext) at (1.6, 0) {Extract};
  \node[stage] (chk) at (3.6, 0) {Chunk};
  \node[stage] (emb) at (5.6, 0) {Embed};
  \node[stage] (idx) at (7.6, 0) {Index};
  \node[io]    (db)  at (9.4, 0) {FAISS+BM25};

  \draw[arr] (pdf) -- (ext);
  \draw[arr] (ext) -- (chk);
  \draw[arr] (chk) -- (emb);
  \draw[arr] (emb) -- (idx);
  \draw[arr] (idx) -- (db);

  \node[anchor=north, font=\tiny\bfseries, color=blue!60!black]
    at (ext.south) {Phase 1};
  \node[anchor=north, font=\tiny\bfseries, color=blue!60!black]
    at (emb.south) {Phase 2 + 3};
  \node[anchor=south west, font=\scriptsize\bfseries]
    at (-0.5, 0.55) {Cold ingestion (once per textbook)};

  % Per query row
  \node[io]    (q)   at (0, -2.4)   {Q};
  \node[stage] (qe)  at (1.6, -2.4) {Embed};
  \node[stage] (sr)  at (3.6, -2.4) {Search};
  \node[stage] (rer) at (5.6, -2.4) {Rerank};
  \node[stage] (gen) at (7.6, -2.4) {Generate};
  \node[io]    (a)   at (9.4, -2.4) {Answer};

  \draw[arr] (q)   -- (qe);
  \draw[arr] (qe)  -- (sr);
  \draw[arr] (sr)  -- (rer);
  \draw[arr] (rer) -- (gen);
  \draw[arr] (gen) -- (a);

  \node[anchor=south west, font=\scriptsize\bfseries]
    at (-0.5, -1.85) {Per query};
\end{tikzpicture}

\vspace{1ex}
\small
\textbf{Phase 1:} faster PDF extraction.
\textbf{Phase 2:} hardware-aware embedding.
\textbf{Phase 3:} pluggable embedder backends.

\note{Show the diagram, point at where each phase sits. Cold path runs
once per textbook, per-query path runs every question. We attack only
the cold path.}
\end{frame}

%% ----------------------------------------------------------
\begin{frame}{What we changed}
\textbf{Phase 1: Faster PDF extraction.}
Parallelize \texttt{docling} across CPU workers via a process pool.
Then add a \texttt{pypdfium2} fast path that skips layout analysis on
plain-text textbooks and recovers section headings with a regex.

\medskip

\textbf{Phase 2: Hardware-aware embedding.}
Sort batches by length so \texttt{llama.cpp}'s padding does not waste
compute. Detect if a GPU is available and pick sequential GPU vs CPU
process pool accordingly.

\medskip

\textbf{Phase 3: Pluggable embedder backends.}
One keyword argument switches between \texttt{llama.cpp} (4\,B GGUF),
GPT4All (Vulkan or CUDA), and HuggingFace
\texttt{sentence-transformers} (Nomic-v1.5, MiniLM).

\note{Each phase is a small surgical change. Phase 1 makes the slow
thing parallel, then replaces the slow thing entirely. Phase 2 cleans
up wasted compute. Phase 3 makes it easy to try lighter embedders.}
\end{frame}

%% ----------------------------------------------------------
\begin{frame}{Results}

\begin{block}{Cold time-to-first-token}
\centering
\textbf{$\sim$\,2h 48m} (original) $\;\longrightarrow\;$
\textbf{$\sim$\,45 s} (4-core / 32\,GiB laptop with 8\,GiB RTX GPU)
\end{block}

\medskip

\textbf{Quality on the 11-question benchmark suite (RTX 6000, 8\,GiB cap):}

\begin{center}
\small
\begin{tabular}{lrrr}
\toprule
variant            & params  & peak VRAM  & final score \\
\midrule
4\,B baseline       & 4\,B    & 5.4\,GiB   & 0.592 \\
\textbf{MiniLM}     & 23\,M   & 2.2\,GiB   & \textbf{0.578} \\
Nomic (ST)          & 137\,M  & 2.7\,GiB   & 0.563 \\
Nomic (GPT4All Vulkan) & 137\,M & 2.1\,GiB & 0.558 \\
\bottomrule
\end{tabular}
\end{center}

\medskip

The 23\,M MiniLM is within 1.4 points of the 4\,B baseline on final
score. Every variant fits comfortably under the 8\,GiB cap. The 4\,B
baseline still runs on the consumer GPU at 5.4\,GiB peak.

\note{Headline number first: nearly three orders of magnitude on TTFT.
Then the table. Don't read every row, just point at MiniLM line and
say it matches the 4B model.}
\end{frame}

%% ----------------------------------------------------------
\begin{frame}{Takeaways}
\begin{itemize}
  \item For plain-text textbooks, layout-aware PDF extraction was
    overkill. \texttt{pypdfium2} is roughly 1{,}100$\times$ faster
    than the original sequential \texttt{docling} extractor and slightly
    \emph{better} on end-to-end answer quality.
  \medskip
  \item On a small corpus, the cross-encoder reranker absorbs most of
    the embedding-quality gap. A 23\,M MiniLM matches a 4\,B model on
    final answer quality, because the right chunk only has to land in
    the FAISS top-50 for the reranker to find it.
  \medskip
  \item TokenSmith now runs end to end on a 4-core / 8\,GiB consumer
    machine. The remaining latency floor is cold extraction, which we
    drove to 8.66\,s on the textbook.
\end{itemize}

\bigskip
\centering
\small
Code: \texttt{github.com/singhals163/TokenSmith}

\note{End on the cross-encoder insight. The story is not about
inventing new tricks; it's about removing the parts that do not earn
their cost.}
\end{frame}

%% ----------------------------------------------------------
\begin{frame}
  \centering
  {\Huge Questions?}\\[2em]
  \small \texttt{github.com/singhals163/TokenSmith}
\end{frame}

\end{document}
