%%% TokenSmith optimization report
%%% ACM SIG Proceedings template (acmart, sigconf option)
%%% Switch to "manuscript, screen, review" for double-spaced submission.

\documentclass[sigconf]{acmart}

\usepackage{booktabs}
\usepackage{xcolor}

%% Disable ACM rights statement / copyright footer for student-project use.
\setcopyright{none}
\acmConference[CS 6423]{CS 6423 Course Project}{Spring 2026}{Georgia Tech}
\acmDOI{}
\acmISBN{}
\settopmatter{printacmref=false, printccs=false, printfolios=true}

\begin{document}

\title{Optimizing TokenSmith for Consumer-GPU Hardware}
\subtitle{A Phase-by-Phase Latency Study of a Local-First Textbook RAG System}

\author{Shivam Singhal}
\email{shivamsinghal16032003@gmail.com}
\affiliation{%
  \institution{Georgia Institute of Technology}
  \city{Atlanta}
  \state{GA}
  \country{USA}
}

\renewcommand{\shortauthors}{Singhal}

\begin{abstract}
TokenSmith is a local-first retrieval-augmented generation (RAG) system
for textbook question answering. It is designed to run on a student's
own machine, with no external API calls. The original implementation
targets server-class GPUs; we re-target it for consumer-class hardware
with at most 8\,GiB of GPU memory and four CPU cores.

We make four pipeline changes. Phase 1 parallelizes PDF extraction
across CPU workers using a process pool, reaching a 25.4$\times$
speedup on a 64-core host and a 3.9$\times$ speedup on four cores.
Phase 1.5 adds a \texttt{pypdfium2} fast extraction path that runs
roughly 57$\times$ faster than \texttt{docling} on plain-text textbooks
without losing section structure. Phase 2 routes embedding work between
a GPU sequential path and a CPU process pool based on hardware
detection, and sorts batches by length to cut padding waste in
\texttt{llama.cpp}. Phase 3 refactors the embedder around a pluggable
backend so the indexer can target \texttt{llama.cpp} GGUF, GPT4All
(including a Vulkan path that needs no CUDA), or HuggingFace
\texttt{sentence-transformers} without changing call sites.

On a Quadro RTX 6000 with peak VRAM capped at 8\,GiB, every embedder
variant stays well below the cap; the largest peak is 5.4\,GiB for the
4\,B baseline. A 23\,M MiniLM variant matches the 4\,B baseline on
end-to-end answer quality (final score 0.578 vs.\ 0.592) while running
13$\times$ faster end to end (40\,s vs.\ 558\,s warm wall).
\end{abstract}

\keywords{retrieval-augmented generation, local inference, embeddings,
GGUF, consumer GPU, parallel PDF extraction, FAISS}

\maketitle

%% ------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

Most RAG systems for textbook Q\&A assume a server with a large GPU
and sustained internet access \cite{lewis2020rag, gao2023rag-survey}.
Students typically do not have either. TokenSmith is a local-first RAG
system that targets a student's own machine: PDFs in, answers out, no
data leaves the host.

Our optimization target in this work is a single laptop-class profile:
one GPU with at most 8\,GiB of memory, four CPU cores, and 32\,GiB of
host RAM. We use the Silberschatz \emph{Database System Concepts}
textbook as the canonical corpus (2{,}195 pages, 2{,}664 chunks after
section-aware splitting). The end-to-end latency target is conservative:
make the answer come back fast enough that a student does not give up
on it during a study session.

The system has four pipeline stages: PDF extraction to markdown, chunk
indexing into FAISS \cite{johnson2019faiss} and BM25
\cite{robertson2009bm25}, retrieval with a cross-encoder reranker, and
generation through a 1.5\,B \texttt{Qwen2.5-Instruct} GGUF model
\cite{qwen2024} running on \texttt{llama.cpp} \cite{llamacpp}. We make
four changes, one per pipeline stage, and measure the impact on a
production server (A100 / V100) and on a Quadro RTX 6000
(Turing, sm\_75) used as a consumer-GPU proxy. Sections~\ref{sec:phase1}
through \ref{sec:phase3} describe each change. Section~\ref{sec:eval}
reports the consumer-GPU evaluation.

%% ------------------------------------------------------------------
\section{System Overview}
\label{sec:system}

The original TokenSmith pipeline runs strictly sequentially.
\texttt{docling} \cite{docling} converts the PDF to markdown one page at
a time, the markdown is chunked on numbered section headings, every
chunk is embedded by a 4\,B \texttt{Qwen3-Embedding} model loaded
through \texttt{llama.cpp}, and the embeddings populate a flat FAISS
index alongside a BM25 sketch. At query time the top-50 candidates are
re-scored by a MiniLM cross-encoder, the top five are passed to the
generator, and a small metric suite (semantic similarity, keyword
overlap, BLEU) compares the answer to a hand-written reference.

The default scoring weights are 0.5 for semantic similarity, 0.3 for
keyword overlap, and 0.2 for BLEU. We refer to the resulting weighted
mean as the \emph{final score}. All numbers in this paper are means
over an 11-question benchmark suite.

\textbf{Hardware referenced in this paper.} The A100 / V100 server
hosts have 64 CPU cores and 80 / 16\,GiB of GPU memory. The RTX 6000
node is a PACE Phoenix allocation with 1$\times$Quadro RTX 6000 (sm\_75,
24\,GiB), four CPU cores, and 32\,GiB of host RAM.

%% ------------------------------------------------------------------
\section{Phase 1: Parallel PDF Extraction}
\label{sec:phase1}

The original extractor calls \texttt{docling} on the entire PDF in one
pass. Layout analysis is the slowest stage of \texttt{docling}, and it
is independent across pages, so we make it parallel.

\subsection{Implementation}

\texttt{split\_pdf} divides the master PDF into fixed-size page-range
chunks and writes each chunk to a temporary file using
\texttt{pypdf} \cite{pypdf}. \texttt{process\_pdf\_chunk} is the
worker: it imports \texttt{docling}, converts one chunk, and serializes
it page by page. A \texttt{ProcessPoolExecutor} runs the workers,
results come back keyed by their starting page offset, and the reduce
phase concatenates them in order. The \texttt{docling} import is
deferred into the worker function so that downstream callers of this
module (the indexer, the pytest harness) do not pay a 100\,MB
shared-library cold-load cost when they only need the markdown
section helpers.

\subsection{Measured Impact}

Table~\ref{tab:phase1-scaling} shows the wall-time scaling. Speedup
saturates near 32 workers on the server host because the
\texttt{pypdf} split phase becomes the Amdahl tail. On a 4-core RTX
6000 host the speedup is bounded by core count and reaches 3.9$\times$
relative to a fresh 4-core sequential run, matching the 4-core
checkpoint baseline reported earlier in the project.

\begin{table}[t]
\caption{Phase 1: docling extraction wall time on the 2{,}195-page
Silberschatz textbook.}
\label{tab:phase1-scaling}
\begin{tabular}{lrrr}
\toprule
host (cores) & workers & wall (s) & speedup \\
\midrule
server (64) & 1  & $\sim$1{,}930 & 1.0$\times$ \\
server (64) & 32 & $\sim$76      & \textbf{25.4$\times$} \\
RTX 6000 (4) & 4 & 499           & 3.9$\times$ \\
\bottomrule
\end{tabular}
\end{table}

%% ------------------------------------------------------------------
\section{Phase 1.5: pypdfium2 Fast Path}
\label{sec:phase15}

Even fully parallel, \texttt{docling} is overkill for plain-text
textbooks. Layout analysis costs minutes per pass and the textbook does
not need it. Phase 1.5 adds a second extraction backend that runs
without layout analysis.

\subsection{Implementation}

The new module \texttt{src/preprocessing/pypdfium\_extractor.py} uses
\texttt{pypdfium2} \cite{pypdfium2} to read raw text from each page
range. Workers seek into the master PDF directly; there is no temporary
split and no map-phase write to disk. A short post-processor strips
table-of-contents lines and promotes lines that look like
``\texttt{N.N Title}'' to ``\texttt{\#\# N.N Title}'' so the existing
markdown section splitter still works. We keep the \texttt{docling}
backend available for documents that need real layout analysis.

\subsection{Measured Impact}

On the RTX 6000 host, full extraction of the 2{,}195-page textbook on
four workers drops from 499\,s with \texttt{docling} to 8.66\,s with
\texttt{pypdfium2}, a $57\times$ improvement. End-to-end final score on
the regenerated index is 0.606, slightly above the 0.592 reported by
the \texttt{docling} pipeline at the same configuration; section
boundaries are recovered well enough that retrieval quality does not
degrade.

%% ------------------------------------------------------------------
\section{Phase 2: Hardware Router and Batch Sorting}
\label{sec:phase2}

Two small changes inside the embedder reduce ingestion wall time on
both server and consumer hardware.

\subsection{Length-Sorted Batches}

\texttt{llama.cpp} pads each embedding batch to the longest text in
that batch. Random ordering wastes compute on short texts that happen
to share a batch with one long text. Before batching we sort the
inputs by length descending; after encoding we permute the result back
to caller order so the FAISS index ID sequence is unchanged. This is a
ten-line change in \texttt{SentenceTransformer.encode()} and removes
the worst case of mixed-length batching.

\subsection{Dynamic GPU / CPU Routing}

\texttt{detect\_gpu()} checks for \texttt{nvidia-smi} (CUDA) or for
Apple Silicon (Metal). When a GPU is reachable the indexer takes a
sequential path with \texttt{batch\_size=32} so the GPU's single
allocator is not contended. When no GPU is reachable the indexer falls
back to a CPU \texttt{multiprocessing.Pool} and lets workers share the
\texttt{mmap}'d GGUF weights. This avoids the worst case where a CPU
worker pool fights the GPU's allocator on a partial-GPU host.

\subsection{Measured Impact}

\begin{table}[t]
\caption{Phase 2: Step 2 (embedding) wall time on 2{,}664 chunks. The
8\,GiB cap is the laptop simulation; \texttt{llama.cpp} is not bound by
torch's allocator and the cap was not binding at any point.}
\label{tab:phase2}
\begin{tabular}{lrr}
\toprule
host & wall (s) & peak VRAM \\
\midrule
A100 (64-core, 80\,GiB)            & 305 & not recorded \\
V100 (64-core, 16\,GiB)            & 426 & 4.3\,GiB \\
RTX 6000 (4-core), uncapped, warm  & \textbf{33.7} & 3.4\,GiB \\
RTX 6000, \texttt{VRAM\_CAP\_GIB=8} & 33.7 & 3.4\,GiB \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:phase2} shows the warm wall-time on each platform. The
RTX 6000 number is the second run in the same shell; the cold start
adds roughly 17\,s for paging in the GGUF weights from Lustre. Peak
VRAM stays at 3.4\,GiB regardless of the cap setting, because
\texttt{llama.cpp} is not bound by torch's per-process memory fraction.

%% ------------------------------------------------------------------
\section{Phase 3: Pluggable Embedding Backends}
\label{sec:phase3}

A 4\,B embedder is heavy. Phase 3 makes the embedder swappable so we
can compare lighter alternatives at a fixed call site.

\subsection{Backend Abstraction}

\texttt{src/embedder.py} introduces a \texttt{BaseEmbedder} ABC with
three concrete implementations: \texttt{LlamaCppEmbedder} for GGUF
models loaded through \texttt{llama.cpp}, \texttt{GPT4AllEmbedder} for
Nomic GGUF models loaded through \texttt{gpt4all} \cite{gpt4all} (with
optional \texttt{device=kompute} for Vulkan and \texttt{device=cuda} for
NVIDIA), and \texttt{HFSentenceTransformerEmbedder} for HuggingFace
\texttt{sentence-transformers} \cite{reimers2019sbert}. A factory
\texttt{make\_embedder(model\_path, backend)} returns the right
implementation. A back-compat alias keeps existing import sites working
without change.

\subsection{Adjacent Fixes}

Phase 3 also adds a sentence-level BLEU-4 metric \cite{papineni2002bleu}
under \texttt{tests/metrics/bleu.py} with NIST method-1 smoothing. We
also fix a pollution bug in \texttt{tests/metrics/semantic.py}: the
original code set \texttt{CUDA\_VISIBLE\_DEVICES=''} at module import,
which silently disabled the GPU for any other process in the same
interpreter. We replace that with \texttt{device='cpu'} on the scoring
model, which restores GPU access for the generator. End-to-end pytest
wall on a typical V100 run drops from $\sim$700\,s to $\sim$45\,s.

\subsection{Quality Comparison on V100}

\begin{table}[t]
\caption{Phase 3, V100, 11-question benchmark suite. Final score is
the weighted mean of semantic, keyword, and BLEU.}
\label{tab:phase3-v100}
\begin{tabular}{lrrrrr}
\toprule
variant & params & sem & kw & BLEU & final \\
\midrule
qwen4b           & 4\,B   & 0.761 & 0.310 & 0.026 & 0.592 \\
nomic\_st        & 137\,M & 0.735 & 0.346 & 0.032 & 0.589 \\
\textbf{minilm}  & 23\,M  & 0.778 & 0.310 & 0.026 & \textbf{0.603} \\
nomic\_gpt4all   & 137\,M & 0.716 & 0.295 & 0.026 & 0.558 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:phase3-v100} reports the four variants. The 23\,M MiniLM
variant slightly beats the 4\,B baseline on final score even though its
retrieval-only quality is clearly weaker. The cross-encoder reranker
recovers most of the gap, as long as the right chunk is somewhere in
the top-50 FAISS neighbours; with a small textbook corpus this
condition holds for most queries.

%% ------------------------------------------------------------------
\section{Consumer-GPU Evaluation}
\label{sec:eval}

The headline question is whether the system runs end to end on a
consumer GPU at 8\,GiB. We answer it on the RTX 6000 with a hard cap.

\subsection{Method}

\texttt{tools/cap\_vram.py} calls
\texttt{torch.cuda.set\_per\_process\_memory\_fraction} based on the
\texttt{VRAM\_CAP\_GIB} environment variable. We set
\texttt{VRAM\_CAP\_GIB=8} for the laptop simulation. \texttt{llama.cpp}
is not bound by the torch allocator, so for the GGUF-backed variant we
report the natural peak VRAM and confirm it stays under 8\,GiB on its
own. We re-run the full pytest suite per variant and report warm-cache
wall time.

\subsection{Results}

\begin{table}[t]
\caption{Phase 3, RTX 6000, \texttt{VRAM\_CAP\_GIB=8}. Final scores
match the uncapped run because the cap is non-binding for every
variant.}
\label{tab:rtx-capped}
\begin{tabular}{lrrr}
\toprule
variant            & wall (s) & peak VRAM (MiB) & final \\
\midrule
minilm             & \textbf{40.3}  & 2{,}214 & 0.578 \\
nomic\_st          & 36.2  & 2{,}706 & 0.563 \\
qwen4b             & 37.1  & 5{,}392 & 0.592 \\
nomic\_gpt4all\_gpu (kompute) & 36.2 & \textbf{2{,}124} & 0.558 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:rtx-capped} shows the four variants under the 8\,GiB
cap. The largest peak is 5.4\,GiB for the 4\,B baseline; every other
variant uses less than 3\,GiB. Final scores are identical to the
uncapped run because the cap only reserves VRAM and does not change
numerics. The Vulkan/kompute path used by GPT4All has the smallest
footprint of the four, and provides a CUDA-free option for non-NVIDIA
silicon.

\subsection{Headline}

The 23\,M MiniLM variant runs end to end in 40 seconds on a 4-core /
8-GiB consumer-GPU profile and matches the 4\,B baseline within one
percentage point on final score. The 4\,B baseline still runs, with
5.4\,GiB peak VRAM, and remains the highest-quality choice for users
who can spare the time. The system is not bottlenecked on memory at
this profile; it is bottlenecked on extraction wall time during cold
ingestion, which Phase 1.5 addresses.

%% ------------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}

\textbf{Why a small embedder is enough on this corpus.} A 23\,M
embedder produces noisier vectors than a 4\,B one, but the difference
is recoverable when the FAISS top-50 still contains the right chunk in
most queries. The cross-encoder reranker re-scores those 50 candidates
on the actual question text and pushes the right one into the top
five. On a much larger corpus this argument would weaken; the cost of
recall@50 grows linearly with corpus size while the reranker cost is
fixed.

\textbf{Why pypdfium2 is enough on this corpus.} Plain-text textbooks
have predictable section headings (\texttt{N.N Title}) and few tables.
A regex post-processor recovers the same chunk boundaries as
\texttt{docling}'s layout pass. On layout-heavy material (papers with
multi-column equations, scanned PDFs), \texttt{docling} would still be
the right default.

\textbf{Threats to validity.} The 11-question benchmark suite is small
and hand-written; the absolute final-score numbers should be read as
relative comparisons rather than absolute quality. The
\texttt{ideal\_retrieved\_chunks} field in the original
\texttt{benchmarks.yaml} was calibrated against an older chunker
schema; we report MiniLM cosine similarity proxies for retrieval-only
quality on top of the keyword-coverage metric.

%% ------------------------------------------------------------------
\section{Software Artifacts}
\label{sec:artifacts}

The work is organized as one branch per feature on the project's git
remote: \texttt{feat/parallel-extraction}, \texttt{feat/pypdfium-extractor},
\texttt{feat/embedder-router}, \texttt{feat/pluggable-embedders}, plus a
\texttt{tools} branch for the resource monitor and the experiment
drivers (profiler, VRAM cap helper, per-variant index/pytest scripts),
and an \texttt{experiments} branch that integrates all five and
deposits per-experiment artifacts in chronological order. Each
\texttt{feat/*} branch is a small surgical delta on \texttt{main} and
is independently mergeable.

%% ------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

Four targeted changes bring TokenSmith from a server-class RAG system
to one that runs in 40 seconds on a 4-core / 8-GiB profile without
losing answer quality. The headline lesson is that the cross-encoder
reranker absorbs most of the embedding-quality gap on small corpora,
which makes lightweight backends genuinely usable for local-first RAG.
The remaining latency floor is cold extraction; Phase 1.5 cuts that to
a few seconds on plain-text textbooks. We did not need new algorithms;
we needed honest measurement and four small changes at the right
seams.

%% ------------------------------------------------------------------
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\end{document}
